---
output: 
  github_document:
always_allow_html: true
fontsize: 12pt
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run:
     Rscript -e 'knitr::knit("README.Rmd")'  -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(knitr)
library(kableExtra)
```

# `walkboutr`

<!-- badges: start -->
<!-- badges: end -->

`walkboutr` turns GPS and accelerometry data into walking bouts. It returns either the
original dataset with bout labels and categories attached, or a summarized, de-identified
dataset that can be shared for collaboration.

## Running the pipeline

The pipeline is **sourced, not installed**. There is no package to build and no
`install.packages()` step: `load.R` reads everything in `R/` into your session.

### In Docker (recommended)

The image is based on `rocker/geospatial`, which ships the GDAL, GEOS and PROJ libraries
that `sf`, `lwgeom` and `geosphere` need already built.

```bash
git clone git@github.com:Population-Dynamics-Lab/walkboutr.git
cd walkboutr

make docker-build     # or: docker build -t walkboutr -f docker/Dockerfile .
make smoke            # runs every step on generated sample data
```

To work interactively, with the repository mounted at `/work`:

```bash
make docker-run       # opens an R session in the image
```

### On your own machine

You need R 4.3 or later and the system libraries for `sf`. Then:

From the repository root:

```r
source("load.R")
```

That is all `smoke.R` and every example below does. `load.R` lists the packages the
pipeline needs and tells you which are missing, with the `install.packages()` call to
fix it.

## How the pipeline works

Three steps run in order. The files in `R/` are numbered to match, so reading them top to
bottom follows the data.

| | File | In | Out |
|---|---|---|---|
| **Step 1** | `R/step1_process_accelerometry_counts_into_bouts.R` | accelerometry counts | counts + `bout`, `non_wearing`, `complete_day` |
| **Step 2** | `R/step2_process_gps_data_into_gps_epochs.R` | GPS fixes | GPS snapped to the epoch grid |
| **Step 3** | `R/step3_process_bouts_and_gps_epochs_into_walkbouts.R` | both of the above | one row per epoch with `bout_category` |

Steps 1 and 2 are independent; step 3 needs both. **Everything else in `R/` is
supporting code with no position in the sequence** — validators, the two data simulators
and the plotting function, each called from the step that needs it. Nothing runs when a
file is sourced, so load order never matters.

`R/pipeline.R` holds the two entry points that run the steps for you. Start there, then
read [`R/README.md`](R/README.md) for the full map.

Settings live in `R/parameters.R`, in two lists: `parameters`, which you can override per
call, and `constants`, which you cannot.

## Basic usage

### Simulated sample data

Simulated data that `walkboutr` can process. The GPS data have the required columns time,
latitude, longitude and speed; the accelerometry data have time and activity counts.
Neither has extra columns, NAs, negative speeds or negative counts, and all times are
date-times in UTC.

```{r simulated data}
source("load.R")
# generate sample gps data:
gps_data <- generate_walking_in_seattle_gps_data()
# generate sample accelerometry data:
accelerometry_counts <- make_full_day_bout_without_metadata()
```

GPS data: 
```{r simulated data: gps, echo = FALSE}
kable(head(gps_data), booktabs = TRUE)  %>% 
  kable_styling(latex_options = "striped", full_width = TRUE, font_size = 12, position = "center")
```

Accelerometry data:
```{r simulated data: acc, echo = FALSE, fig.align='center'}
kable(head(accelerometry_counts), booktabs = TRUE)  %>% 
  kable_styling(latex_options = "striped", full_width = TRUE, font_size = 12, position = "center")
```

<p>

There are two top level functions, giving either (1) a dataset with bouts and bout
categories alongside all of your original data, or (2) a summary dataset that is
de-identified and shareable.

### Walk bout dataset including original data

```{r example 1, message = FALSE, warning = FALSE, results = 'hide'}
walk_bouts <- identify_walk_bouts_in_gps_and_accelerometry_data(gps_data, accelerometry_counts)
```

```{r example printout, echo = FALSE}
kable(head(walk_bouts)) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::kable_styling(font_size = 12) %>% 
  kableExtra::kable_styling(position = "center")
```

### Summarized walk bout dataset 

A set of labelled bouts, categorized (`bout_category`), with each bout's median speed
(`median_speed`), start time (`bout_start`), duration in minutes (`duration`), and a flag
for whether the bout came from a complete day of data (`complete_day`).

```{r example 2, message = FALSE, warning = FALSE}
summary <- summarize_walk_bouts(walk_bouts)
```

```{r example 2 printout, echo = FALSE}
kable(head(summary)) %>% 
  kableExtra::kable_styling(latex_options = "striped", font_size = 12, position = "center")
summary_1 <- summary %>% dplyr::filter(bout == 1)
```

This example has `r nrow(summary)` bout(s), each with a label. Bout 1 started at
`r format(summary_1$bout_start, format = "%Y-%m-%d %H:%M:%S")`, lasted
`r round(summary_1$duration, 1)` minutes, and was categorized as
`r summary_1$bout_category`. The calendar day it falls on
`r ifelse(summary_1$complete_day, "has", "does not have")` enough wearing time to count as
complete.

Bouts are categorized in precedence order, highest first: `non_walk_incomplete_gps`,
`dwell_bout`, `non_walk_too_vigorous`, `non_walk_slow`, `non_walk_fast`. A bout matching
none of those is a `walk_bout`. For more detail see the **Generate Walk Bouts** vignette.

## Repository layout

```
R/            the pipeline, numbered by step; see R/README.md
load.R        checks dependencies and sources R/ into your session
smoke.R       runs every step on sample data and prints the result
docker/       the runtime image
Makefile      docker-build, docker-run, smoke, lint
docs/         longer walkthroughs, formerly the package vignettes
paper.md      the article describing the method
```

## Further reading

- [`R/README.md`](R/README.md) — what each file does and the order to read them in
- [`docs/process_bouts.Rmd`](docs/process_bouts.Rmd) — bout categories explained in full
- [`docs/changing_default_parameters.Rmd`](docs/changing_default_parameters.Rmd) — the
  parameters you can override
- [`docs/generate_data.Rmd`](docs/generate_data.Rmd) — the simulated data generators
- `paper.md` — the article describing the method

## Note on the R package

`walkboutr` was published to CRAN as an R package version 0.5.0. This repository is no
longer an R package: `DESCRIPTION`, `NAMESPACE`, `man/`, `tests/` and the pkgdown
configuration have been removed, and the vignettes now live in `docs/` as ordinary
documents. The roxygen `#'` comments in `R/` are kept as documentation, but their tags
are inert since nothing processes them.

See `NEWS.md` for version 0.6.0, which fixes bugs that change bout classification
relative to the published 0.5.0.
