---
output: 
  github_document:
    html_preview: false
fontsize: 12pt
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run:
     Rscript render_readme.R   (or: make readme)  -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
library(knitr)
```

# `walkboutr`

<!-- badges: start -->
<!-- badges: end -->

`walkboutr` turns GPS and accelerometry data into walking bouts. It returns either the
original dataset with bout labels and categories attached, or a summarized, de-identified
dataset that can be shared for collaboration.

## Running the pipeline

The pipeline is **sourced, not installed**. There is no package to build and no
`install.packages()` step: `load.R` reads everything in `R/` into your session.

### With repo2docker and RStudio (recommended)

This is the standard PDL path. It builds an image with R, RStudio and every dependency,
then hands you a browser-based RStudio session.

```bash
repo2docker https://github.com/Population-Dynamics-Lab/walkboutr
```

Copy the link the console prints, open it, then **View > Open Jupyter Lab > RStudio**.
In the RStudio console:

```r
source("load.R")    # loads the pipeline
source("smoke.R")   # runs every step and prints the result
```

`runtime.txt` pins R 4.4.1 and a Posit Package Manager snapshot date, so package versions
are fixed. `install.R` lists what gets installed.

If you are on a Mac with Apple silicon, install OrbStack first and grant it command line
access, then `uv tool install jupyter-repo2docker`.

### With the plain Dockerfile

A smaller, command-line-only image, with no RStudio or Jupyter. Based on
`rocker/geospatial`, which ships the GDAL, GEOS and PROJ libraries that `sf`, `lwgeom` and
`geosphere` need already built.

```bash
git clone git@github.com:Population-Dynamics-Lab/walkboutr.git
cd walkboutr

make docker-build     # or: docker build -t walkboutr -f docker/Dockerfile .
make smoke            # runs every step on generated sample data
make docker-run       # an interactive R session, repo mounted at /work
```

### On your own machine

You need R 4.3 or later and the system libraries for `sf`. From the repository root:

```r
source("load.R")
```

That is all `smoke.R` and every example below does. `load.R` lists the packages the
pipeline needs and tells you which are missing, with the `install.packages()` call to
fix it.

## `R/` — what runs in what order

**Only three files are steps, and they are numbered.** Everything else is supporting
code called *from* those steps. There is no order to the rest, and there does not need
to be: every file here only defines functions, except `parameters.R`, which defines two
lists. Nothing executes when a file is sourced, so `load.R` can read them in any order.

If you are reading the code for the first time, go in this order:

| Order | File | Role |
|---|---|---|
| 1 | `pipeline.R` | **Start here.** The two entry points. Shows the three steps in sequence in about five lines. |
| 2 | `parameters.R` | Every threshold the pipeline uses, in two lists. Read before the steps so the numbers in them mean something. |
| 3 | `step1_process_accelerometry_counts_into_bouts.R` | Accelerometry counts → candidate bouts. |
| 4 | `step2_process_gps_data_into_gps_epochs.R` | GPS fixes → epoch-aligned GPS. Independent of step 1. |
| 5 | `step3_process_bouts_and_gps_epochs_into_walkbouts.R` | Both of the above → categorised walk bouts. |

### Execution order

```
                accelerometry_counts        gps_data
                         |                     |
                    [ step 1 ]            [ step 2 ]
                         |                     |
                       bouts              gps_epochs
                         \                    /
                          \                  /
                           +--> [ step 3 ] <-+
                                     |
                              walk_bouts (categorised)
                                     |
                          summarize_walk_bouts()
                                     |
                         one row per bout, de-identified
```

Steps 1 and 2 are independent and could run in either order. Step 3 needs both.

### The supporting files

These have no position in the sequence. Each is called from the step that needs it.

| File | Called by | What it does |
|---|---|---|
| `parameters.R` | all steps | `parameters` (overridable per call), `constants` (not overridable), and `collate_arguments()`, which merges caller overrides in. |
| `validate_accelerometry_data.R` | step 1 | Rejects malformed accelerometry input before anything else runs. |
| `validate_gps_data.R` | step 2 | The same for GPS input. |
| `plot.R` | you, directly | The three article figures, one function each. |
| `simulate_gps_data.R` | examples, `smoke.R` | Generates GPS data for a walk in Seattle. |
| `simulate_accelerometry_data.R` | examples, `smoke.R` | The `make_*` builders, each constructing a specific scenario (smallest bout, full day, non-wearing period). |
| `simulate_example_bouts.R` | the figures | Data shaped like real data, for plotting. The `make_*` builders above produce square waves and `generate_gps_data()` wanders for miles, neither of which makes a readable figure. |

### Bout categories

Assigned in `step3`. Every bout starts as `walk_bout` and each rule overwrites the
previous one, so the **last** rule to match wins. In precedence order, highest first:

1. `non_walk_incomplete_gps` — too few GPS records, or too little coverage
2. `dwell_bout` — the bounding circle is smaller than the 66 ft threshold
3. `non_walk_too_vigorous` — mean counts above the walking ceiling
4. `non_walk_slow` — median speed below the walking floor
5. `non_walk_fast` — median speed above the walking ceiling
6. `walk_bout` — matched none of the above

## Basic usage

### Simulated sample data

Simulated data that `walkboutr` can process. The GPS data have the required columns time,
latitude, longitude and speed; the accelerometry data have time and activity counts.
Neither has extra columns, NAs, negative speeds or negative counts, and all times are
date-times in UTC.

```{r simulated data}
source("load.R")
# generate sample gps data:
gps_data <- generate_walking_in_seattle_gps_data()
# generate sample accelerometry data:
accelerometry_counts <- make_full_day_bout_without_metadata()
```

GPS data: 
```{r simulated data: gps, echo = FALSE}
kable(head(gps_data))
```

Accelerometry data:
```{r simulated data: acc, echo = FALSE, fig.align='center'}
kable(head(accelerometry_counts))
```

<p>

There are two top level functions, giving either (1) a dataset with bouts and bout
categories alongside all of your original data, or (2) a summary dataset that is
de-identified and shareable.

### Walk bout dataset including original data

```{r example 1, message = FALSE, warning = FALSE, results = 'hide'}
walk_bouts <- identify_walk_bouts_in_gps_and_accelerometry_data(gps_data, accelerometry_counts)
```

```{r example printout, echo = FALSE}
kable(head(walk_bouts))
```

### Summarized walk bout dataset 

A set of labelled bouts, categorized (`bout_category`), with each bout's median speed
(`median_speed`), start time (`bout_start`), duration in minutes (`duration`), and a flag
for whether the bout came from a complete day of data (`complete_day`).

```{r example 2, message = FALSE, warning = FALSE}
summary <- summarize_walk_bouts(walk_bouts)
```

```{r example 2 printout, echo = FALSE}
kable(head(summary))
summary_1 <- summary %>% dplyr::filter(bout == 1)
```

This example has `r nrow(summary)` bout(s), each with a label. Bout 1 started at
`r format(summary_1$bout_start, format = "%Y-%m-%d %H:%M:%S")`, lasted
`r round(summary_1$duration, 1)` minutes, and was categorized as
`r summary_1$bout_category`. The calendar day it falls on
`r ifelse(summary_1$complete_day, "has", "does not have")` enough wearing time to count as
complete.

Bouts are categorized in precedence order, highest first: `non_walk_incomplete_gps`,
`dwell_bout`, `non_walk_too_vigorous`, `non_walk_slow`, `non_walk_fast`. A bout matching
none of those is a `walk_bout`. For more detail see the **Generate Walk Bouts** vignette.

## Repository layout

```
R/              the pipeline, numbered by step
load.R          checks dependencies and sources R/ into your session
smoke.R         runs every step on sample data and prints the result
runtime.txt     R version and CRAN snapshot date, read by repo2docker
install.R       packages repo2docker installs at build time
docker/         an alternative command-line-only image
Makefile        docker-build, docker-run, smoke, lint, readme
render_readme.R regenerates README.md from README.Rmd
docs/           longer walkthroughs, formerly the package vignettes
paper.md        the article describing the method
```

`README.Rmd` is the only README you edit. It has live R chunks, so the tables and
bout counts above are computed rather than typed. `make readme` renders it to
`README.md`, which is what GitHub shows. Editing `README.md` is pointless; the next
render overwrites it.

## Reproducing the article's figures

Each of the three figures in `paper.md` is one function call, so a script that
regenerates them needs no plotting code of its own:

```{r figures, eval = FALSE}
example_data <- make_dwell_and_walk_example()

generate_activity_bout_plot(make_activity_bout_example(), bout_number = 1)
generate_dwell_vs_walk_plot(example_data$accelerometry_counts, example_data$gps_data)
generate_bout_plot(example_data$accelerometry_counts, example_data$gps_data, bout_number = 2)
```

Each takes raw accelerometry counts and GPS data and runs whatever pipeline steps it
needs. `make_dwell_and_walk_example()` produces a matched pair of bouts whose bounding
circles fall either side of the 66 ft dwell threshold, which is what the second figure
needs in order to show anything.

## Further reading

- [`docs/process_bouts.Rmd`](docs/process_bouts.Rmd) — bout categories explained in full
- [`docs/changing_default_parameters.Rmd`](docs/changing_default_parameters.Rmd) — the
  parameters you can override
- [`docs/generate_data.Rmd`](docs/generate_data.Rmd) — the simulated data generators
- `paper.md` — the article describing the method

## Note on the R package

`walkboutr` was published to CRAN as an R package version 0.5.0. This repository is no
longer an R package: `DESCRIPTION`, `NAMESPACE`, `man/`, `tests/` and the pkgdown
configuration have been removed, and the vignettes now live in `docs/` as ordinary
documents. The roxygen `#'` comments in `R/` are kept as documentation, but their tags
are inert since nothing processes them.

See `NEWS.md` for version 0.6.0, which fixes bugs that change bout classification
relative to the published 0.5.0.
